---
title: "Estimands: ATT, ATE, ATC"
author: "Jens Hainmueller"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Estimands: ATT, ATE, ATC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 96
)
set.seed(20260505)
```

`ebalance()` reweights one or both groups to match a target moment
distribution. Which target you pick determines the **estimand** — the
average treatment effect on a particular subpopulation. As of 0.3-0,
`ebalance(..., estimand = "ATT" / "ATE" / "ATC")` selects:

| `estimand` | reweighted | target moments | answers |
|---|---|---|---|
| `"ATT"` (default) | controls | treated group means | "what was the effect on those who actually got treatment?" |
| `"ATC"` | treated   | control group means | "what would the effect have been on the control population if it had been treated?" |
| `"ATE"` | both      | overall sample means | "what is the average effect across the whole population?" |

This vignette builds the same toy panel and shows what changes across
the three estimands.

## A toy panel

```{r setup}
library(ebal)
set.seed(20260505)
n0 <- 200; n1 <- 100
X <- rbind(
  replicate(3, rnorm(n0, mean = 0)),       # controls
  replicate(3, rnorm(n1, mean = 0.5))      # treated, shifted
)
colnames(X) <- c("x1", "x2", "x3")
treatment <- c(rep(0, n0), rep(1, n1))

c(treated_mean = colMeans(X[treatment == 1, ])[1],
  control_mean = colMeans(X[treatment == 0, ])[1],
  overall_mean = mean(X[, 1]))
```

The treated group means are above zero; the controls are at zero; the
overall mean is somewhere between.

## Three fits

```{r fits}
fit_att <- ebalance(Treatment = treatment, X = X, estimand = "ATT")
fit_ate <- ebalance(Treatment = treatment, X = X, estimand = "ATE")
fit_atc <- ebalance(Treatment = treatment, X = X, estimand = "ATC")
```

`weights(fit)` returns a length-`n` vector aligned to the original
treatment indicator. The shape changes with the estimand:

```{r weights-shape}
table(treatment, sign(weights(fit_att)))   # ATT: treated = 1, controls reweighted
table(treatment, sign(weights(fit_ate)))   # ATE: both reweighted
table(treatment, sign(weights(fit_atc)))   # ATC: treated reweighted, controls = 1
```

## Where the weight goes

For ATT, the controls are reweighted toward the treated mean (so their
weighted mean equals the treated mean):

```{r att-balance}
weighted.mean(X[treatment == 0, 1], w = weights(fit_att)[treatment == 0])
mean(X[treatment == 1, 1])
```

For ATE, **both groups** are reweighted toward the overall mean:

```{r ate-balance}
weighted.mean(X[treatment == 0, 1], w = weights(fit_ate)[treatment == 0])
weighted.mean(X[treatment == 1, 1], w = weights(fit_ate)[treatment == 1])
mean(X[, 1])
```

For ATC, the treated are reweighted toward the control mean:

```{r atc-balance}
weighted.mean(X[treatment == 1, 1], w = weights(fit_atc)[treatment == 1])
mean(X[treatment == 0, 1])
```

## Diagnostics

`glance()` returns a one-row "is this fit usable?" summary, with
per-side ESS / max-weight ratios and the worst pre/post standardized
difference:

```{r glance, eval = requireNamespace("generics", quietly = TRUE)}
library(generics)
do.call(rbind, lapply(list(ATT = fit_att, ATE = fit_ate, ATC = fit_atc),
                      glance))[, c("estimand", "ess_control", "ess_treated",
                                   "max_weight_ratio_control",
                                   "max_weight_ratio_treated",
                                   "max_abs_std_diff_post")]
```

Read these top-down: ATT keeps the treated side trivial (ESS = 100,
ratio = 1) and concentrates weight on a subset of controls. ATE
reweights both sides, so both ESS values fall below their group sizes.
ATC mirrors ATT with roles swapped.

## Plots

`autoplot(fit, type = "balance")` is the Love plot of standardized
differences before vs. after weighting. `autoplot(fit, type = "weights")`
is a histogram of the per-unit weights with the Kish ESS and
max-weight ratio in the subtitle:

```{r autoplot, eval = requireNamespace("ggplot2", quietly = TRUE)}
library(ggplot2)
autoplot(fit_ate, type = "weights")
```

Identical interfaces are available in base graphics via
`plot(fit, type = ...)`.

## Choosing an estimand

A short rule of thumb:

- **ATT** when the policy question is about people who actually got
  treated (most program-evaluation work). Controls are a tool for
  imputing the counterfactual; you don't need to extrapolate to them.
- **ATE** when you want a population-level claim ("if everyone got
  this treatment...") and you trust that the treatment effect is
  similar across the covariate distribution.
- **ATC** when the policy question is about the controls
  ("would extending this program to non-recipients have helped?"); be
  honest about the extrapolation involved.

`ebalance()` doesn't make this choice for you — it just gives you the
weights once you've made it. The associated standard error / inference
question is independent (see `vignette("outcome-models", package = "ebal")`).
